---
title: "Oil in marine protected areas"
subtitle: "Which MPAs are exposed to satellite-detected oil slicks, and does protection level matter?"
date: today
format:
  html:
    theme: none
    css: template/custom.css
    include-in-header: template/header.html
    include-after-body: template/footer.html
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 9
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(data.table)
library(ggplot2)
library(scales)
library(reactable)
library(sf)
source("R/plots.R")

exposure <- as.data.table(readRDS("data/mpa_exposure.rds"))
cls      <- as.data.frame(readRDS("data/class_summary.rds"))
pts      <- as.data.table(readRDS("data/slick_points.rds"))
info     <- readRDS("data/run_info.rds")

fmt  <- function(x, d = 0) format(round(x, d), big.mark = ",", nsmall = d)
pct  <- function(x) sprintf("%.0f%%", 100 * x)

oiled   <- exposure[n_slicks > 0]
by_cls  <- exposure[, .(n = .N, oiled = sum(n_slicks > 0),
                        share = mean(n_slicks > 0)), by = .(prot_class, class_label)][order(prot_class)]
top     <- as.data.table(top_mpas(as.data.frame(exposure)))
a       <- cls[cls$class == "all", ]
```

## The question

In September 2026 SkyTruth published a global estimate of marine oil pollution:
around 1.6 million km² of ocean surface covered by oil slicks every year, most of
it from ships and offshore platforms, and most of it inside national waters. Their
report notes in passing that marine protected areas are not spared.

This note takes that further with the same underlying data, SkyTruth's
[Cerulean](https://cerulean.skytruth.org) slick detections from Sentinel-1
radar, and asks two questions:

1. **Which MPAs are actually exposed to oil**, by name, and how much?
2. **Does the level of protection make a difference**, once you account for
   where MPAs sit and how often satellites look at them?

Between `r format(info$start, "%B %Y")` and `r format(info$end, "%B %Y")`,
Cerulean recorded `r fmt(info$n_slicks)` slicks matched to a nearby vessel or
platform. Of the `r fmt(nrow(exposure))` marine protected areas in the WDPA,
**`r fmt(nrow(oiled))` (`r pct(nrow(oiled) / nrow(exposure))`) had at least one
detected inside them.**

## The most exposed MPAs

```{r top, fig.height = 9}
plot_top_mpas(as.data.frame(exposure), info)
```

Ranking by oil area per unit of MPA area favours small coastal sites near busy
shipping lanes and ports, which is the point: these are the places where a
protected designation and chronic oil exposure coincide most intensely. MPAs
smaller than 100 km² or with fewer than three slicks are left out, because a
single large slick in a tiny reserve produces a spectacular density that says
little about chronic exposure.

The top of the list is
**`r top$NAME[1]`** (`r top$country[1]`, `r top$class_label[1]`), with
`r top$n_slicks[1]` slicks covering `r fmt(top$oil_km2[1], 1)` km² inside its
boundary over `r info$n_years` years.

### The largest totals

Density hides the big offshore MPAs, where exposure is diluted by sheer size but
the absolute amount of oil can be large.

```{r total}
tt <- exposure[n_slicks > 0][order(-oil_km2)][1:15]
tt[, y := sprintf("%s  (%s)", ifelse(nchar(NAME) > 42, paste0(substr(NAME, 1, 40), "\u2026"), NAME), country)]
tt[, y := factor(y, levels = rev(unique(y)))]

ggplot(tt, aes(oil_km2, y, fill = as.character(prot_class))) +
  geom_col(width = 0.65) +
  geom_text(aes(label = sprintf("%s slicks", fmt(n_slicks))), hjust = -0.15,
            size = 3, colour = INK2) +
  scale_fill_manual(values = CLASS_COLS, labels = info$class_labels,
                    breaks = names(info$class_labels)) +
  scale_x_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.2))) +
  coord_cartesian(clip = "off") +
  labs(title = "Where the most oil ends up inside MPAs",
       subtitle = sprintf("Total oil-slick area detected inside each MPA, %s-%s",
                          format(info$start, "%Y"), format(info$end, "%Y")),
       x = "Oil-slick area inside the MPA (km²)", y = NULL) +
  theme_oil()
```

## Where

```{r map, fig.height = 5.5}
world <- rnaturalearth::ne_countries(scale = 110, returnclass = "sf")
m <- st_as_sf(oiled, coords = c("lon", "lat"), crs = 4326)
bg <- st_as_sf(pts[sample(.N, min(.N, 30000))], coords = c("lon", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = world, fill = "#e9e8e3", colour = NA) +
  geom_sf(data = bg, colour = "#52514e", size = 0.05, alpha = 0.25) +
  geom_sf(data = m[order(-m$oil_km2), ], aes(size = oil_km2, fill = as.character(prot_class)),
          shape = 21, colour = SURF, stroke = 0.3, alpha = 0.9) +
  scale_size_area(max_size = 9, labels = label_comma(), name = "km² of oil") +
  scale_fill_manual(values = CLASS_COLS, labels = info$class_labels,
                    breaks = names(info$class_labels), name = NULL) +
  coord_sf(crs = "EPSG:8857", expand = FALSE) +
  labs(title = "MPAs with oil detected inside them",
       subtitle = "Circles: MPAs, sized by oil-slick area inside. Grey dots: all detected slicks (sample).") +
  theme_oil() +
  guides(size = guide_legend(override.aes = list(fill = MUTED)),
         fill = guide_legend(override.aes = list(size = 4))) +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank(),
        panel.grid.major = element_blank(), legend.box = "vertical")
```

The pattern follows SkyTruth's global hotspots: the Mediterranean and Black
Sea, the Red Sea and Persian Gulf, West Africa and Southeast Asia. The grey dots
also show the limit of this kind of data: open-ocean areas are sparse because
Sentinel-1 rarely images them, not necessarily because they are clean.

## Does protection level matter?

Comparing the oil density inside MPAs with the global ocean average would be
misleading: MPAs are mostly coastal, where both shipping and satellite coverage
are concentrated. Instead, each protection class is compared with the
unprotected ocean within `r info$ring_km` km of it, which shares the same
satellite revisit pattern, shipping context and sea state.

```{r classes, fig.height = 7}
plot_classes(cls, info)
```

Across all MPAs, oil density inside is `r sprintf("%.2f", a$ratio)` times that of
the surrounding ocean (95% CI `r sprintf("%.2f-%.2f", a$ratio_lo, a$ratio_hi)`).

```{r class-table}
reactable(
  data.table(`Protection class` = cls$label,
             MPAs = cls$n_mpas,
             `With slicks` = cls$n_mpas_oiled,
             `Share with slicks` = sprintf("%.1f%%", 100 * cls$n_mpas_oiled / cls$n_mpas),
             `Density inside` = round(cls$dens_in, 3),
             `Density around` = round(cls$dens_ring, 3),
             `Ratio [95% CI]` = sprintf("%.2f [%.2f-%.2f]", cls$ratio, cls$ratio_lo, cls$ratio_hi)),
  compact = TRUE, pagination = FALSE,
  defaultColDef = colDef(format = colFormat(separators = TRUE)))
```

Densities are km² of slick per 10,000 km² of ocean per year.

## The full table

All MPAs with at least one slick detected inside them.

```{r full}
ft <- oiled[order(-oil_per_1000km2_yr),
            .(MPA = NAME, Country = country, Designation = DESIG_ENG,
              IUCN = IUCN_CAT, Class = class_label,
              `Area (km²)` = round(marine_km2),
              Slicks = n_slicks, `Months with slicks` = n_months,
              `Oil inside (km²)` = round(oil_km2, 2),
              `Oil per 1,000 km² per yr` = round(oil_per_1000km2_yr, 2))]
reactable(ft, searchable = TRUE, defaultPageSize = 15, compact = TRUE, highlight = TRUE,
          defaultColDef = colDef(format = colFormat(separators = TRUE)))
```

## Caveats

**Detection is not measurement.** Cerulean detects slicks on Sentinel-1 radar
images. A place imaged twice a month will record fewer slicks than one imaged
every few days, whatever the true pollution. The per-MPA figures are therefore
*detected* exposure, and the open ocean and some high-latitude regions are
under-sampled. The class comparison controls for this with neighbouring waters;
the per-MPA ranking does not.

**Only source-matched slicks.** Following SkyTruth's recommendation, slicks are
kept only if Cerulean's attribution model links them to a nearby vessel or
offshore platform (collated score > 0). This removes most false positives but
also slicks from unknown sources, which biases the sample towards shipping lanes
and oil fields.

**Protection class is taken from IUCN management categories as reported to the
WDPA.** Categories I-III are treated as no-take, IV-VI as multiple-use. Many MPAs
report no category at all, and the category describes management intent, not
enforcement.

**Slick area is the surface footprint on the day of the image,** not a volume.
A thin sheen and a thick spill of the same extent count the same.

**Attribution is proximity, not proof.** A slick inside an MPA was detected
there; it was not necessarily released there, and the nearby vessel is a
candidate, not a culprit.

## Reproducing this

```bash
Rscript prep/build_data.R      # downloads + analysis, writes data/ (30-60 min first run)
Rscript linkedin_figures.R     # PNGs in outputs/
quarto render index.qmd
```

Slicks: `r info$slick_filter`, `r format(info$start)` to `r format(info$end)`.
Grid: `r info$res_km` km equal-area (Equal Earth). Ring: `r info$ring_km` km.
Bootstrap: `r fmt(info$n_boot)` resamples of months. Data built on
`r format(info$run_date)`.

Oil slick data: SkyTruth Cerulean, derived from Copernicus Sentinel-1 imagery.
Protected areas: UNEP-WCMC and IUCN, World Database on Protected Areas,
[protectedplanet.net](https://www.protectedplanet.net). Coastlines: Natural Earth.
