---
title: "Oil in marine protected areas"
subtitle: "Which countries and which MPAs are exposed to satellite-detected oil slicks?"
date: today
format:
  html:
    theme: none
    css: template/custom.css
    include-in-header: template/header.html
    include-after-body: template/footer.html
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 9
    fig-height: 6
    lightbox: true          # click any figure to open it full screen
    code-fold: true         # code shown collapsed under each chunk
    code-summary: "Show code"
    code-tools: true
    highlight-style: breezedark     # pandoc built-in dark theme (works on any Quarto version)
knitr:
  opts_chunk:
    dev.args:
      bg: "#052832"       # figure background = page background (no white edges)
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| code-summary: "Setup: packages and data"
library(data.table)
library(ggplot2)
library(scales)
library(reactable)
library(sf)
source("R/plots.R")
use_dark_theme()   # charts and tables match the dark report stylesheet

exposure <- as.data.table(readRDS("data/mpa_exposure.rds"))
cls      <- as.data.frame(readRDS("data/class_summary.rds"))
pts      <- as.data.table(readRDS("data/slick_points.rds"))
info     <- readRDS("data/run_info.rds")

fmt  <- function(x, d = 0) format(round(x, d), big.mark = ",", nsmall = d)
pct  <- function(x) sprintf("%.0f%%", 100 * x)

oiled   <- exposure[n_slicks > 0]
by_cls  <- exposure[, .(n = .N, oiled = sum(n_slicks > 0),
                        share = mean(n_slicks > 0)), by = .(prot_class, class_label)][order(prot_class)]
top     <- as.data.table(top_mpas(as.data.frame(exposure)))
a       <- cls[cls$class == "all", ]
```

## The question

In September 2026 SkyTruth published a global estimate of marine oil pollution:
around 1.6 million km² of ocean surface covered by oil slicks every year, most of
it from ships and offshore platforms, and most of it inside national waters. Their
report notes in passing that marine protected areas are not spared.

This note takes that further with the same underlying data, SkyTruth's
[Cerulean](https://cerulean.skytruth.org) slick detections from Sentinel-1
radar, and asks which countries and which protected areas are actually exposed
to oil, and how much.

Between `r format(info$start, "%B %Y")` and `r format(info$end, "%B %Y")`,
Cerulean recorded `r fmt(info$n_slicks)` slicks matched to a nearby vessel or
platform. Of the `r fmt(nrow(exposure))` marine protected areas in the WDPA,
**`r fmt(nrow(oiled))` (`r pct(nrow(oiled) / nrow(exposure))`) had at least one
detected inside them.**

## Country by country

Which countries have the most oil inside their protected areas, and is it a
few sites or most of them? Each square below is one MPA; countries are ranked
by the total oil-slick area detected inside their MPAs.

```{r country-ranking, fig.width = 8, fig.height = 8}
#| code-summary: "Show code: country ranking"
plot_country_ranking(as.data.frame(exposure), info, 20)
```

The squares show how that oil is spread across each country's network: a few
heavily hit sites, or most of its MPAs.

```{r countries, fig.width = 10, fig.height = country_grid_height(as.data.frame(exposure), 20)}
#| code-summary: "Show code: country grid"
plot_country_grid(as.data.frame(exposure), info, n_countries = 20)
```

## MPA-level exposure

Oil exposure summarised by IUCN protection level, then MPA by MPA. Both tables
are sortable (click a column header) and the second one is searchable.

### By protection level

```{r class-table}
#| code-summary: "Show code: protection-level table"
cls_tbl <- data.frame(
  `Protection class` = cls$label,
  MPAs               = cls$n_mpas,
  `With slicks`      = cls$n_mpas_oiled,
  `Share with slicks` = 100 * cls$n_mpas_oiled / cls$n_mpas,
  `Density inside`   = cls$dens_in,
  `Density around`   = cls$dens_ring,
  `Ratio [95% CI]`   = sprintf("%.2f [%.2f-%.2f]", cls$ratio, cls$ratio_lo, cls$ratio_hi),
  check.names = FALSE)

reactable(cls_tbl, theme = table_theme(), compact = TRUE, pagination = FALSE,
          resizable = TRUE,                 # drag a header's right edge to resize
          striped = TRUE, highlight = TRUE,
          columns = list(
            `Protection class`  = colDef(minWidth = 190, style = list(fontWeight = 600)),
            `Share with slicks` = colDef(format = colFormat(suffix = "%", digits = 1)),
            `Density inside`    = colDef(format = colFormat(digits = 3)),
            `Density around`    = colDef(format = colFormat(digits = 3))),
          defaultColDef = colDef(format = colFormat(separators = TRUE)))
```

Densities are km² of oil slick per 10,000 km² of ocean per year. "Around" is the
unprotected ocean within `r info$ring_km` km of the MPAs, which shares their
satellite coverage and shipping context, so the ratio compares like with like.

### All MPAs with oil

Every MPA with at least one slick detected inside it, ordered by the total
oil-slick area inside its boundary (the same measure as the country chart).
"Oil per 1,000 km²" divides that by the MPA's size: it puts small, heavily hit
reserves at the top when you sort by it, which is why it is not the default order.

```{r full}
#| code-summary: "Show code: MPA table"
ft <- oiled[order(-oil_km2),
            .(MPA = NAME, Country = country, Designation = DESIG_ENG,
              IUCN = IUCN_CAT, `Area (km²)` = marine_km2,
              Slicks = n_slicks, `Months with slicks` = n_months,
              `Oil inside (km²)` = oil_km2,
              `Oil per 1,000 km² per yr` = oil_per_1000km2_yr)]
lvl <- poll_bins(ft$`Oil per 1,000 km² per yr`)
heat <- setNames(POLL_COLS_LIGHT[seq_along(levels(lvl))], levels(lvl))

# Colour key for the shaded column (light table: darker = more oil)
key <- htmltools::div(
  style = "display:flex;flex-wrap:wrap;gap:.35rem 1rem;align-items:center;font-size:.8rem;color:#9cb7c9;margin:.2rem 0 .6rem;",
  htmltools::span(style = "color:#ffffff;font-weight:600;", "Oil per 1,000 km\u00b2 per yr:"),
  lapply(names(heat), function(l) htmltools::span(
    style = "display:inline-flex;align-items:center;gap:.35rem;",
    htmltools::span(style = sprintf(
      "width:.95rem;height:.95rem;border-radius:2px;display:inline-block;background:%s;", heat[[l]])),
    l)),
  htmltools::span(style = "font-style:italic;", "darker = more oil"))

tbl <- reactable(ft, theme = table_theme(), searchable = TRUE, defaultPageSize = 25,
          resizable = TRUE,                 # drag a header's right edge to resize
          compact = TRUE, striped = TRUE, highlight = TRUE,
          height = 640,                     # scroll inside the table: header stays put
          columns = list(
            # MPA and Country stay frozen on the left when scrolling sideways
            MPA         = colDef(minWidth = 220, sticky = "left",
                                 style = list(fontWeight = 600, background = "inherit"),
                                 headerStyle = list(background = "#cfe0e6")),
            Country     = colDef(minWidth = 120, sticky = "left",
                                 style = list(background = "inherit",
                                              borderRight = "1px solid #b7cdd5"),
                                 headerStyle = list(background = "#cfe0e6",
                                                    borderRight = "1px solid #b7cdd5")),
            Designation = colDef(minWidth = 160),
            IUCN        = colDef(minWidth = 115, align = "center"),
            `Area (km²)` = colDef(format = colFormat(separators = TRUE, digits = 0)),
            `Oil inside (km²)` = colDef(format = colFormat(separators = TRUE, digits = 1)),
            `Oil per 1,000 km² per yr` = colDef(
              format = colFormat(digits = 2),
              style = function(value, index) {
                bg <- heat[[as.character(lvl[index])]]
                list(background = bg, fontWeight = 600,
                     color = if (as.integer(lvl[index]) >= 4) "#ffffff" else TABLE_INK)
              })),
          defaultColDef = colDef(format = colFormat(separators = TRUE)))

htmltools::tagList(key, tbl)
```

## Caveats

**Detection is not measurement.** Cerulean detects slicks on Sentinel-1 radar
images. A place imaged twice a month will record fewer slicks than one imaged
every few days, whatever the true pollution. The per-MPA figures are therefore
*detected* exposure, and the open ocean and some high-latitude regions are
under-sampled. The class comparison controls for this with neighbouring waters;
the per-MPA ranking does not.

**Only source-matched slicks.** Following SkyTruth's recommendation, slicks are
kept only if Cerulean's attribution model links them to a nearby vessel or
offshore platform (collated score > 0). This removes most false positives but
also slicks from unknown sources, which biases the sample towards shipping lanes
and oil fields.

**Protection class is taken from IUCN management categories as reported to the
WDPA.** Categories I-III are treated as no-take, IV-VI as multiple-use. Many MPAs
report no category at all, and the category describes management intent, not
enforcement.

**Slick area is the surface footprint on the day of the image,** not a volume.
A thin sheen and a thick spill of the same extent count the same.

**Attribution is proximity, not proof.** A slick inside an MPA was detected
there; it was not necessarily released there, and the nearby vessel is a
candidate, not a culprit.

## Reproducing this

```bash
Rscript prep/build_data.R      # downloads + analysis, writes data/ (30-60 min first run)
Rscript linkedin_figures.R     # PNGs in outputs/
quarto render index.qmd
```

Slicks: `r info$slick_filter`, `r format(info$start)` to `r format(info$end)`.
Grid: `r info$res_km` km equal-area (Equal Earth). Ring: `r info$ring_km` km.
Bootstrap: `r fmt(info$n_boot)` resamples of months. Data built on
`r format(info$run_date)`.

The chart functions live in `R/plots.R`:

```{r plots-code, eval = FALSE, file = "R/plots.R"}
#| code-summary: "Show code: R/plots.R"
```

Oil slick data: SkyTruth Cerulean, derived from Copernicus Sentinel-1 imagery.
Protected areas: UNEP-WCMC and IUCN, World Database on Protected Areas,
[protectedplanet.net](https://www.protectedplanet.net). Coastlines: Natural Earth.
